
```{r include=FALSE}
library(rbytecode)
```

# Exploration {#sec-exploration}

This chapter explores some of the inner workings of R and the R bytecode compiler.

## Compiler Optimization: Constant folding {#sec-constant-folding}

The internal R bytecode compiler includes *constant folding* as one of 
its optimization mechanisms.

[Constant folding](https://en.wikipedia.org/wiki/Constant_folding) is the 
process of evaluating constant expressions at compile time rather than computing
them at execution time.

Constants include simple literals (e.g. `2`), but also include expressions whose
value is known at compile time.

There are four types of objects that are **not** treated as constants:

1. function calls of type *language*
2. variable references of type *symbol*
3. promises
4. bytecode objects

If an expression does not contain any of these, then it is a candidate for
evaluation during the bytecode compilation process i.e. *constant folding*.

According to Tierney [@tierney23] *constant folding* is applied during code 
generation, and it not part of a separate optimization phase.

Note: `{rbytecode}` bypasses R's compilation mechanism and does not include
any optimization steps i.e. bytecode assembly is converted to a bytecode object
exactly as specified.

### Simple demonstration of *constant folding* in action

As an example of constant folding, compare the disassembly of the following
two expressions `1 + x` and `1 + 2`

```{r}
disq({
  1 + x
}) |> as.character()
```

In `x + 1`, the variable `x` is not a constant (i.e. it is a variable
reference of type *symbol* - which is the second in the list of four non-constant
elements noted above).  Since the expression contains a non-constant, it is left
as-is and evaluated at runtime i.e. the value of `x` is always fetched in order to 
calculate the result.

```{r}
disq({
  1 + 2
}) |> as.character()
```

In `1 + 2`, both arguments are constants and the compiler evaluates this
expression at compile time.  When evaluating this bytecode, it simply fetches
the pre-computed value of `3` and returns it.

Note that the function `+` does *not* count as a function call of type *language*
in this example.  Because `+` is a *primitive* operation it gets a free-pass and
its calculation is assumed to be deterministic.


### More complex example with constant folding

Constant folding works with numeric vectors and nested operations.

In the following, the actual compiled bytecode simply loads the pre-calculated
answer and returns it.

```{r}
disq(
  1:3 + 2^2 * (4 + (8 - 2))
) |> as.character()
```



### Constant Folding and the Commutative Property

In the following examples it is evident that in R's compiler, the optimization step related
to constant folding does not make use of the [commutative property](https://en.wikipedia.org/wiki/Commutative_property).

`2 * 3 * x` is correctly optimized to `6 * x`.

But in `2 * x * 3`, no simplification is done and the disassembled bytecode indicates that 
`2 * x` is evaluated first, and then that result is multiplied by 3.

```{r}
disq(
  2 * 3 * x
) |> as.character()
```


```{r}
disq(
  2 * x * 3
) |> as.character()
```


### Constant folding and `MATH1` operations

It is interesting to note that even though the `MATH1` functions ([Section -@sec-math1functions]) 
are deterministic, they are not included in *constant folding* optimization.

In the following example, `sin(1)` always evaluates to the same value, yet
the bytecode indicates that this value is calculated every time during execution
by calling the `sin()` function with an argument of `1`.

```{r}
disq(
  sin(1)
) |> as.character()
```



### Corner case: redefined primitive functions

What happens if `+` is not a primitive operation?  This would be the case if
its definition had been changed for some reason.

```{r}
`+` <- function(x, y) {x + y}

disq(
  1 + 2
) |> as.character()
```

Now that the compiler detects this function is no longer the primitive it 
usually is, it can no longer determine that the expression always
evaluates to a constant.   No constant folding is performed, and the bytecode 
will call the function each time it is evaluated.

### Future ideas

It is possible that future optimization steps in the R compiler could address:

1. Consider `MATH1` functions with constant arguments as producing a constant
    result which can be optimized with *constant folding*.
2. Apply the commutative property to find more opportunities for *constant folding*.




